---
title: "Chapter 6 Empirical Examples"
author: "Xianwen He"
date: '2022-07-12'
output: 
  prettydoc::html_pretty:
    toc: True
    theme: hpstr
    highlight: github
---

## Reisenzein's example (1986)

This section aims to reproduce to empirical examples in Chapter 6, Section 6.7. Please install `MIIVsem` and `lavaan` using `install.packages('<package.name>')` if you haven't done so.

Load the packages.
```{r, warning=F, message=F}
library(MIIVsem)
library(lavaan)
```

Read in the data.
```{r}
reisenzein1986 <- readRDS("reisenzein1986a.RDS")
colnames(reisenzein1986) <- c(
"Z1", "Z2", "Z3", "Z4", "Z5", "Z6",
"Z7", "Z8", "Z9", "Z10", "Z11", "Z12",
"Z13"
)
```

### The original model ($M_1$)

Specify the original model ($M_1$).
```{r}
m1.reisenzein <- '
L1 =~ Z1 + Z2 + Z3
L2 =~ Z4 + Z5 + Z6
L3 =~ Z7 + Z8 + Z9
L4 =~ Z10 + Z11 + Z12
L4 ~ L2 + L3
L2 ~ L1
L3 ~ L1
L1 ~ Z13
'
```

ML fit measures using `lavaan`.
```{r}
fitmeasures(sem(m1.reisenzein, data = reisenzein1986))
```

2SLS estimates using `MIIVsem`.
```{r}
miive(m1.reisenzein, reisenzein1986, sarg.adjust= "holm")
```

### $M_2$

Specify the model. Compared to $M_1$, there is an direct impact from $Z_{13}$('Stimulus') to $L_4$('Help'), and there are tow pairs of empirically determined correlated errors ($Z_3$ & $Z_4$, $Z_4$ & $Z_{12}$).

```{r}
m2.reisenzein <- '
  L1 =~ Z1 + Z2 + Z3
  L2 =~ Z4 + Z5 + Z6
  L3 =~ Z7 + Z8 + Z9
  L4 =~ Z10 + Z11 + Z12
 
  L4 ~ L2 + L3 + Z13
  L2 ~ L1
  L3 ~ L1
  L1 ~ Z13
 
  Z3 ~~ Z4
  Z4 ~~ Z12
'
```

ML fit measures for $M_2$.
```{r}
fitmeasures(sem(m2.reisenzein, data = reisenzein1986))
```


### The cross-loading model ($M_3$)

Specify the model. Add a direct path from $L_4$('Help') to $Z_6$ ('Concern').
```{r}
m3.reisenzein <- '
L1 =~ Z1 + Z2 + Z3
L2 =~ Z4 + Z5 + Z6
L3 =~ Z7 + Z8 + Z9
L4 =~ Z10 + Z11 + Z12 + Z6
L4 ~ L2 + L3
L2 ~ L1
L3 ~ L1
L1 ~ Z13
'
```

ML fit measures.
```{r}
fitmeasures(sem(m3.reisenzein, data = reisenzein1986))
```

2SLS estimates for $M_3$.
```{r}
miive(m3.reisenzein, reisenzein1986, sarg.adjust= "holm")
```

## Higher-order factor analysis

This section aims to reproduce the empirical examples in Chapter 6, Section 6.8.1.

Read in the data from the study by Colom et al. (2004).
```{r}
colom.n <- 198
lower.corr <- '
1
0.21	1
0.24	0.11	1
0.34	0.14	0.07	1
0.42	0.12	0.09	0.40	1
0.41	0.19	0.16	0.50  0.49	1
0.19	0.19	0.17	0.18	0.20  0.26	1
0.19	0.22	0.09	0.13	0.04	0.22	0.43	1
0.32	0.17	0.04	0.06	0.16	0.29	0.21	0.39	1
0.30  0.14	0.26	0.20  0.13	0.26	0.28	0.20  0.25	1
0.35	0.23	0.30  0.19	0.06	0.21	0.20  0.32	0.35	0.39	1
0.32	0.29	0.12	0.10  0.17	0.24	0.18	0.42	0.33	0.30  0.49	1
'
mean <- '26.3  30.1  8.9  39.6  37.8  38.3  35  35.2  22.9  15.7  22.4  23.9'
std <- '10.3  9.6  8.6  10.9  9.6  10.5  6.9  6.7  6.5  3.4  10.6  4.6'
labels <- c('z1', 'z2', 'z3', 'z5', 'z6', 'z4',
            'z8', 'z7', 'z9', 'z11', 'z10', 'z12')

colom.corr <- getCov(lower.corr, names=labels)
colom.mean <- as.numeric(strsplit(mean, '  ')[[1]]); names(colom.mean) <- labels
colom.std <- as.numeric(strsplit(std, '  ')[[1]]); names(colom.std) <- labels

# sigma_i * sigma_j * rho_ij = Cov(X_i, X_j)
colom.cov <- (colom.std %*% t(colom.std)) * colom.corr
dimnames(colom.cov) <- list(labels,labels)
```

Define the model.
```{r}
colom.mod <- '
L1 =~ L2 + L3 + L4 + L5
# measurements
L2 =~ z1 + z2 + z3
L3 =~ z4 + z5 + z6
L4 =~ z7 + z8 + z9
L5 =~ z10 + z11 + z12
'
```

ML estimates by `lavaan`.
```{r}
colom.fit <- sem(colom.mod,
    sample.cov = colom.cov, sample.mean = colom.mean,
    sample.nobs = colom.n)
# test statistics
fitmeasures(colom.fit)
# estimates
summary(colom.fit, standardized = TRUE)
```

Specify two MIIVs for each observed variable and applies `MIIVsem` for 2SLS estimates. The script for MIIV selection can be found in the folder supplement.
```{r}
colom.miiv <- '
z4 ~ z9 + z10
z7 ~ z6 + z10
z10 ~ z6 + z9
z2 ~ z3 + z6
z3 ~ z2 + z6
z5 ~ z6 + z1
z6 ~ z5 + z9
z8 ~ z9 + z12
z9 ~ z8 + z12
z11 ~ z12 + z3
z12 ~ z11 + z9
'

miive(model = colom.mod, instruments = colom.miiv,
      sample.cov = colom.cov, sample.mean = colom.mean,
      sample.nobs = colom.n)
```



## Longitudinal model

This section aims to reproduce the empirical examples in Chapter 6, Section 6.8.2. 

Read in the NLSY data.
```{r, warning=F, message=F}
dat.long <- read.table('./halaby2.dat', header=TRUE, sep=" ", dec=",")
```

### Fixed Effects Model ($M_1$)

Fit a FEM without lagged effects.
```{r}
# FEM without lagged effects
m1.fem<-'
#measurement model
alpha=~1*lnwg80+1*lnwg81+1*lnwg82+1*lnwg83+1*lnwg84+1*lnwg85+1*lnwg86+1*lnwg87
lnwg80~1
lnwg81~1
lnwg82~1
lnwg83~1
lnwg84~1
lnwg85~1
lnwg86~1
lnwg87~1
#lnwg80~~v*lnwg80
#lnwg81~~v*lnwg81
#lnwg82~~v*lnwg82
#lnwg83~~v*lnwg83
#lnwg84~~v*lnwg84
#lnwg85~~v*lnwg85
#lnwg86~~v*lnwg86
#lnwg87~~v*lnwg87
#structural model
lnwg80~a1*SEI.10_80+a2*union80
lnwg81~a1*SEI.10_81+a2*union81
lnwg82~a1*SEI.10_82+a2*union82
lnwg83~a1*SEI.10_83+a2*union83
lnwg84~a1*SEI.10_84+a2*union84
lnwg85~a1*SEI.10_85+a2*union85
lnwg86~a1*SEI.10_86+a2*union86
lnwg87~a1*SEI.10_87+a2*union87

alpha~0
alpha~~SEI.10_80+SEI.10_81+SEI.10_82+SEI.10_83+SEI.10_84+SEI.10_85+SEI.10_86+SEI.10_87+union80+union81+union82+union83+union84+union85+union86+union87
SEI.10_80~~SEI.10_81+SEI.10_82+SEI.10_83+SEI.10_84+SEI.10_85+SEI.10_86+SEI.10_87+union80+union81+union82+union83+union84+union85+union86+union87
SEI.10_81~~SEI.10_82+SEI.10_83+SEI.10_84+SEI.10_85+SEI.10_86+SEI.10_87+union80+union81+union82+union83+union84+union85+union86+union87
SEI.10_82~~SEI.10_83+SEI.10_84+SEI.10_85+SEI.10_86+SEI.10_87+union80+union81+union82+union83+union84+union85+union86+union87
SEI.10_83~~SEI.10_84+SEI.10_85+SEI.10_86+SEI.10_87+union80+union81+union82+union83+union84+union85+union86+union87
SEI.10_84~~SEI.10_85+SEI.10_86+SEI.10_87+union80+union81+union82+union83+union84+union85+union86+union87
SEI.10_85~~SEI.10_86+SEI.10_87+union80+union81+union82+union83+union84+union85+union86+union87
SEI.10_86~~SEI.10_87+union80+union81+union82+union83+union84+union85+union86+union87
SEI.10_87~~union80+union81+union82+union83+union84+union85+union86+union87
union80~~union81+union82+union83+union84+union85+union86+union87
union81~~union82+union83+union84+union85+union86+union87
union82~~union83+union84+union85+union86+union87
union83~~union84+union85+union86+union87
union84~~union85+union86+union87
union85~~union86+union87
union86~~union87'

fit.1.fem <- sem(m1.fem, data=dat.long, fixed.x=FALSE)
# summary(fit.1.fem, fit.measure=TRUE)
fitmeasures(fit.1.fem)
```

### Random Effects Model ($M_2$)

Fit a REM without lagged effects.
```{r}
m2.rem<-'
#measurement model
alpha=~1*lnwg80+1*lnwg81+1*lnwg82+1*lnwg83+1*lnwg84+1*lnwg85+1*lnwg86+1*lnwg87
lnwg80~1
lnwg81~1
lnwg82~1
lnwg83~1
lnwg84~1
lnwg85~1
lnwg86~1
lnwg87~1
#lnwg80~~v*lnwg80
#lnwg81~~v*lnwg81
#lnwg82~~v*lnwg82
#lnwg83~~v*lnwg83
#lnwg84~~v*lnwg84
#lnwg85~~v*lnwg85
#lnwg86~~v*lnwg86
#lnwg87~~v*lnwg87
#structural model
lnwg80~a1*SEI.10_80+a2*union80+a3*educ+a4*black
lnwg81~a1*SEI.10_81+a2*union81+a3*educ+a4*black
lnwg82~a1*SEI.10_82+a2*union82+a3*educ+a4*black
lnwg83~a1*SEI.10_83+a2*union83+a3*educ+a4*black
lnwg84~a1*SEI.10_84+a2*union84+a3*educ+a4*black
lnwg85~a1*SEI.10_85+a2*union85+a3*educ+a4*black
lnwg86~a1*SEI.10_86+a2*union86+a3*educ+a4*black
lnwg87~a1*SEI.10_87+a2*union87+a3*educ+a4*black

alpha~0
'

fit.2.rem<-sem(m2.rem, data=dat.long, fixed.x=FALSE)
# summary(fit.2.rem, fit.measure=TRUE)
fitmeasures(fit.2.rem)
```

### Latent Growth Curve Model ($M_3$)

Fit a linear growth model.
```{r}
m3.gr_lin<-'
#measurement model
alpha=~1*lnwg80+1*lnwg81+1*lnwg82+1*lnwg83+1*lnwg84+1*lnwg85+1*lnwg86+1*lnwg87
beta=~0*lnwg80+1*lnwg81+2*lnwg82+3*lnwg83+4*lnwg84+5*lnwg85+6*lnwg86+7*lnwg87

lnwg80~0
lnwg81~0
lnwg82~0
lnwg83~0
lnwg84~0
lnwg85~0
lnwg86~0
lnwg87~0
lnwg80~~lnwg80
lnwg81~~lnwg81
lnwg82~~lnwg82
lnwg83~~lnwg83
lnwg84~~lnwg84
lnwg85~~lnwg85
lnwg86~~lnwg86
lnwg87~~lnwg87
#structural model
lnwg80~a1*SEI.10_80+a2*union80
lnwg81~a1*SEI.10_81+a2*union81
lnwg82~a1*SEI.10_82+a2*union82
lnwg83~a1*SEI.10_83+a2*union83
lnwg84~a1*SEI.10_84+a2*union84
lnwg85~a1*SEI.10_85+a2*union85
lnwg86~a1*SEI.10_86+a2*union86
lnwg87~a1*SEI.10_87+a2*union87

alpha~1
beta~1
alpha~educ+black
beta~educ+black
alpha~~beta'

fit.3.gr_lin<-sem(m3.gr_lin, data=dat.long)
# summary(fit.3.gr_lin, fit.measure=TRUE)
fitmeasures(fit.3.gr_lin)
```

### Autoregressive Model ($M_4$)

Fit a straight auto-regressive model.
```{r}
m4.ar<-'
#measurement model
lnwg80~1
lnwg81~1
lnwg82~1
lnwg83~1
lnwg84~1
lnwg85~1
lnwg86~1
lnwg87~1
lnwg80~~lnwg80
lnwg81~~lnwg81
lnwg82~~lnwg82
lnwg83~~lnwg83
lnwg84~~lnwg84
lnwg85~~lnwg85
lnwg86~~lnwg86
lnwg87~~lnwg87
#structural model
lnwg80~SEI.10_80+union80+a3*educ+a4*black
lnwg81~rho*lnwg80+a1*SEI.10_81+a2*union81+a3*educ+a4*black
lnwg82~rho*lnwg81+a1*SEI.10_82+a2*union82+a3*educ+a4*black
lnwg83~rho*lnwg82+a1*SEI.10_83+a2*union83+a3*educ+a4*black
lnwg84~rho*lnwg83+a1*SEI.10_84+a2*union84+a3*educ+a4*black
lnwg85~rho*lnwg84+a1*SEI.10_85+a2*union85+a3*educ+a4*black
lnwg86~rho*lnwg85+a1*SEI.10_86+a2*union86+a3*educ+a4*black
lnwg87~rho*lnwg86+a1*SEI.10_87+a2*union87+a3*educ+a4*black
'

fit.4.ar<-sem(m4.ar,data=dat.long)
# summary(fit.4.ar, fit.measure=TRUE)
fitmeasures(fit.4.ar)
```

### LV-ALT1 ($M_5$)

Fit a classic ALT model with linear growth and constant auto-regressive parameters.
```{r}
m5.lv_lin<-'
#measurement model
eta_1=~lnwg80
eta_2=~lnwg81
eta_3=~lnwg82
eta_4=~lnwg83
eta_5=~lnwg84
eta_6=~lnwg85
eta_7=~lnwg86
eta_8=~lnwg87
alpha=~1*eta_2+1*eta_3+1*eta_4+1*eta_5+1*eta_6+1*eta_7+1*eta_8
beta=~1*eta_2+2*eta_3+3*eta_4+4*eta_5+5*eta_6+6*eta_7+7*eta_8

lnwg80~0
lnwg81~0
lnwg82~0
lnwg83~0
lnwg84~0
lnwg85~0
lnwg86~0
lnwg87~0
lnwg80~~v1*lnwg80
lnwg81~~v1*lnwg81
lnwg82~~lnwg82
lnwg83~~lnwg83
lnwg84~~lnwg84
lnwg85~~lnwg85
lnwg86~~lnwg86
lnwg87~~lnwg87

#structural model
eta_2~rho2*eta_1+a1*SEI.10_81+a2*union81+a3*educ+a4*black
eta_3~rho3*eta_2+a1*SEI.10_82+a2*union82+a3*educ+a4*black
eta_4~rho4*eta_3+a1*SEI.10_83+a2*union83+a3*educ+a4*black
eta_5~rho5*eta_4+a1*SEI.10_84+a2*union84+a3*educ+a4*black
eta_6~rho6*eta_5+a1*SEI.10_85+a2*union85+a3*educ+a4*black
eta_7~rho7*eta_6+a1*SEI.10_86+a2*union86+a3*educ+a4*black
eta_8~rho8*eta_7+a1*SEI.10_87+a2*union87+a3*educ+a4*black

alpha~1
beta~1
eta_1~1
eta_1~~eta_1
eta_2~~v*eta_2
eta_3~~v*eta_3
eta_4~~v*eta_4
eta_5~~v*eta_5
eta_6~~v*eta_6
eta_7~~v*eta_7
eta_8~~v*eta_8
alpha~~0*black+0*educ
beta~~0*black+0*educ
alpha~~eta_1+beta+ SEI.10_81+SEI.10_82+SEI.10_83+SEI.10_84+SEI.10_85+SEI.10_86+SEI.10_87+union81+union82+union83+union84+union85+union86+union87
beta~~eta_1+SEI.10_81+SEI.10_82+SEI.10_83+SEI.10_84+SEI.10_85+SEI.10_86+SEI.10_87+union81+union82+union83+union84+union85+union86+union87
eta_1~~SEI.10_81+SEI.10_82+SEI.10_83+SEI.10_84+SEI.10_85+SEI.10_86+SEI.10_87+union81+union82+union83+union84+union85+union86+union87+educ+black
SEI.10_81~~SEI.10_82+SEI.10_83+SEI.10_84+SEI.10_85+SEI.10_86+SEI.10_87+union81+union82+union83+union84+union85+union86+union87+educ+black
SEI.10_82~~SEI.10_83+SEI.10_84+SEI.10_85+SEI.10_86+SEI.10_87+union81+union82+union83+union84+union85+union86+union87+educ+black
SEI.10_83~~SEI.10_84+SEI.10_85+SEI.10_86+SEI.10_87+union81+union82+union83+union84+union85+union86+union87+educ+black
SEI.10_84~~SEI.10_85+SEI.10_86+SEI.10_87+union81+union82+union83+union84+union85+union86+union87+educ+black
SEI.10_85~~SEI.10_86+SEI.10_87+union81+union82+union83+union84+union85+union86+union87+educ+black
SEI.10_86~~SEI.10_87+union81+union82+union83+union84+union85+union86+union87+educ+black
SEI.10_87~~union81+union82+union83+union84+union85+union86+union87+educ+black
union81~~union82+union83+union84+union85+union86+union87+educ+black
union82~~union83+union84+union85+union86+union87+educ+black
union83~~union84+union85+union86+union87+educ+black
union84~~union85+union86+union87+educ+black
union85~~union86+union87+educ+black
union86~~union87+educ+black
union87~~educ+black
educ~~black
'

fit.5.lv_lin <- sem(m5.lv_lin, data=dat.long, fixed.x=FALSE)
```


ML fit measures.
```{r}
fitmeasures(fit.5.lv_lin)
```

ML estimates.
```{r}
summary(fit.5.lv_lin, fit.measure=FALSE)
```

### LV-ALT2 ($M_6$ )

Fit a LV-ALT with linear growth and unrestricted auto-regressive parameters.
```{r}
m6.lv_lin_new<-'
#measurement model
eta_1=~lnwg80
eta_2=~lnwg81
eta_3=~lnwg82
eta_4=~lnwg83
eta_5=~lnwg84
eta_6=~lnwg85
eta_7=~lnwg86
eta_8=~lnwg87
alpha=~1*eta_2+1*eta_3+1*eta_4+1*eta_5+1*eta_6+1*eta_7+1*eta_8
beta=~1*eta_2+2*eta_3+3*eta_4+4*eta_5+5*eta_6+6*eta_7+7*eta_8

lnwg80~0
lnwg81~0
lnwg82~0
lnwg83~0
lnwg84~0
lnwg85~0
lnwg86~0
lnwg87~0
lnwg80~~var*lnwg80
lnwg81~~var*lnwg81
lnwg82~~lnwg82
lnwg83~~lnwg83
lnwg84~~lnwg84
lnwg85~~lnwg85
lnwg86~~lnwg86
lnwg87~~lnwg87
#structural model
eta_2~rho*eta_1+a1*SEI.10_81+a2*union81
eta_3~rho*eta_2+a1*SEI.10_82+a2*union82
eta_4~rho*eta_3+a1*SEI.10_83+a2*union83
eta_5~rho*eta_4+a1*SEI.10_84+a2*union84
eta_6~rho*eta_5+a1*SEI.10_85+a2*union85
eta_7~rho*eta_6+a1*SEI.10_86+a2*union86
eta_8~rho*eta_7+a1*SEI.10_87+a2*union87

alpha~1
beta~1
eta_1~1
alpha~educ+black
beta~educ+black
eta_1~SEI.10_80+union80+educ+black
eta_1~~eta_1
eta_2~~v*eta_2
eta_3~~v*eta_3
eta_4~~v*eta_4
eta_5~~v*eta_5
eta_6~~v*eta_6
eta_7~~v*eta_7
eta_8~~v*eta_8
alpha~~eta_1+beta
beta~~eta_1
'

fit.6.lv_lin_new<-sem(m6.lv_lin_new, data=dat.long, fixed.x=FALSE)
```


ML fit measures.
```{r}
fitmeasures(fit.6.lv_lin_new)
```

ML estimates.
```{r}
summary(fit.6.lv_lin_new, fit.measure=FALSE)
```


## Bayes SEM

This section aims to reproduce the empirical examples in Chapter 6, Section 6.9.

Load the package.
```{r, warning=F, message=F}
library("blavaan")
```

Read in the data.
```{r}
dat.bollen <- MIIVsem::bollen1989a
# rename the variables.
colnames(dat.bollen) <- c("Z4", "Z5", "Z6", 
                          "Z7", "Z8", "Z9", 
                          "Z10", "Z11", 
                          "Z1", "Z2", "Z3" )
```

Summarize the dataset.
```{r}
dat.summ <- t(apply(dat.bollen, 2,
                    function(x){
                      return (c(quantile(x), mean(x), sd(x)))
                      }
                    ))
dat.summ <- data.frame(rbind(dat.summ[9:11, ], dat.summ[1:8, ]))
colnames(dat.summ) <- c('Min', '25%', 'Median', '75%', 'Max', 'Mean', 'Sd.')
dat.summ
```



### Estimation

Specify the model.
```{r}
mod.indem <- '
  L1 =~ Z1 + Z2 + Z3 
  L2 =~ Z4 + Z5 + Z6  + Z7  
  L3 =~ Z8 + Z9 + Z10 + Z11    

  L2  ~ L1  
  L3  ~ L1 + L2
 
  L1 ~ 1
  Z1 ~ 0
  L2 ~ 1
  Z4 ~ 0
  L3 ~ 1
  Z8 ~ 0
  
  Z4 ~~ Z8
  Z5 ~~ Z7 + Z9
  Z6 ~~ Z10
  Z7 ~~ Z11
  Z9 ~~ Z11 '
```


MIIV-2SLS estimation with `MIIVsem`.
```{r}
miive(mod.indem, dat.bollen)
```

ML estimation with `lavaan`.
```{r}
Lavfit <- sem(mod.indem, data = dat.bollen)
summary(Lavfit, standardized = TRUE)
```

Bayesian estimation with `blavaan`. 

**Note**: This chunk takes a long time to run. Thus I leave it here simply for display and read in an already generated Bayesian SEM object for the subsequent analysis. Please check the Rmarkdown for details.

```{r, eval=F}
# two chains with 50,000 iterations for each
Bfit <- bsem(mod.indem, data=dat.bollen, bcontrol = list(cores = 2L),
             n.chains = 2, burnin = 50000, sample = 5000)
```

```{r, echo=F}
# This is the line to read in an already generated Bayesian object. 
# You can find the script for saving this checkpoint in the folder supplement.
# Please use your own model to update the RData file if necessary.
# Note that the following outputs may be different
# because of the randomness of the Bayesian estimates.
load('./Bfit.RData')
```


Summarize the Bayesian model. Note that results are unlikely to be exactly the same with those in book because of the randomness.
```{r}
# Huge difference from the results in the draft
summary(Bfit)
```

### Fit measures

Fit a null model to calculate CFI, TLI, and NFI.

**Note**: This chunk also takes a long time to run. I implement the same strategy as above. Please check the Rmarkdown for details.
```{r, eval=F}
# null model
null.model <- c(paste0("Z", 1:11, " ~~ Z", 1:11), paste0("Z", 1:11, " ~ 1"))
fit0 <- bsem(null.model, data = dat.bollen, bcontrol = list(cores = 2L),
             n.chains = 2, burnin = 50000, sample = 5000)
```

```{r, echo=FALSE}
# This is the line to read in an already generated null model.
# Please use your own model to update fit0.RData.
load('fit0.RData')
```

Fit measure for the Bayesian SEM.
```{r, warning=F}
B.ml <- blavFitIndices(Bfit, baseline.model = fit0)
summary(B.ml)
```

### Trace plots

Trace plots for factor loadings and regression parameters.
```{r}
plot(Bfit, c(3:4, 10:11))
```

